pirouette: Determining the error BEAST2 makes in inferring a phylogeny

### 2019-03-04 article.tex
Changes @rsetienne

% Affiliations
\usepackage{authblk}
\title{pirouette: the error BEAST2 makes in inferring a phylogeny}
\author[1]{Rich\`el J.C. Bilderbeek}
\author[1]{Giovanni Laudanno}
\author[1]{Rampal S. Etienne}
\affil[1]{Groningen Institute for Evolutionary Life Sciences, University of Groningen, Groningen, The Netherlands}

@@

%comments
\newcommand{\giovanni}[1]{\textcolor{blue}{\textbf{[GL: #1]}}}
\newcommand{\richel}[1]{\textcolor{orange}{\textbf{[RB: #1]}}}

@@

  \textbf{1. }
    BEAST2 is a popular Bayesian phylogenetics software tool,
    that takes an alignment and an inference model to create a
    posterior of jointly-estimated phylogenies and model parameter estimates.
    When a new macro-evolutionary diversification model is developed,
    a good first step is to measure the error BEAST2 makes, with its current set of inference models, on a known
    phylogeny derived from a new diversification mechanism. \\
  \textbf{2. }
    Here we present \verb;pirouette;, 
    a free, libre and open-source R package that assesses 
    the inference error BEAST2 makes based on a known/true 
    phylogeny. \\
  \textbf{3. }
    We describe \verb;pirouette;'s usage and the biological scientific
    question it can answer, including full examples. \\
  \textbf{4. }
    Last, we will discuss the results obtained by the examples. \\
\end{abstract}

@@

The development of new powerful inference tools, 
such as BEAST [\cite{drummond2007beast}], 
MrBayes [\cite{huelsenbeck2001mrbayes}]
or RevBayes [\cite{hohna2016revbayes}], 
allows to build phylogenetic trees 
from genetic material extracted from extant organisms.
This has constituted an important step forward 
in our understanding on how species evolve.
Such tools have been increasingly exploited to hypothesize 
and test what the main drivers and modes for diversification are.

Since BEAST performs a Bayesian analysis, 
it needs not only genetic data but also tree priors, 
describing how the diversification occur, to return posterior phylogenies.
BEAST works using standard tree priors like Yule or birth-death 
\giovanni{integrate this if necessary}.
The recent development of BEAST2 [\cite{bouckaert2014beast}] provides, 
unlike its predecessor, the possibility for third-party users 
to integrate its prior libraries with new ones. 
Many of these have been developed, 
in the form of likelihood diversification models \giovanni{Not of them are already available for BEAST2 yet, or maybe they will never be}: 
birth-death models that have rates that are 
constant [\cite{nee1994reconstructed}], 
time-dependent [\cite{nee1994reconstructed}, \cite{rabosky2008explosive}], 
diversity-dependent [\cite{etienne2011diversity}],
or can shift [\cite{etienne2012conceptual}, the SLS model. Laudanno et al., in preparation].
There are birth death models in which speciation is a process that takes time [\cite{rosindell2010protracted}][\cite{etienne2012prolonging}], can co-occur [Laudanno et al., in preparation],
dependent on a trait that is binary [\cite{maddison2007estimating}], 
has multiple states [\cite{fitzjohn2012diversitree}],
has a concealed state [\cite{beaulieu2016detecting}] 
or has multiple observed and concealed states [\cite{herrera2018detecting}].

From a given phylogeny, the likelihood is maximized
to infer the best estimates for the parameter values.
For such  models, it is a standard practice to assess their performance 
on a controlled environment: this is done by simulating phylogenies from known parameters, 
then comparing the estimations obtained by maximum likelihood with the original parameters.
If the likelihood proves to be effective to describe the process, 
it can potentially be implemented into BEAST2 as a new prior. 
Despite that, it is still possible that already existing BEAST2 priors 
will perform well enough and, therefore, 
that the introduction of a new model is not necessary.

Diversification models are getting more complex over time. 
This is facilitated by the increase in computing power due to the advancement of technology.
Increased complexity made possible the exploration of new models 
encompassing mechanisms that are not yet considered by previous diversification models. 
An example of this is the multiple-birth model [Laudanno et al., in preparation],
that is the first model to allow simultaneous speciations to co-occur.

However, whenever a new model is proposed there is always a trade-off 
between the cost of increased complexity versus the benefits that it can provide.
In fact developing models can cost time, human and computational resources.
Therefore desirable models have to be able to disentangle 
the intrinsic complexity of a system, 
highlight the main reasons of its behaviour and avoid unnecessary complexity.

For this reason we believe it is important to have techniques 
to quantitatively assess the advancement provided by the introduction of new models, 
in comparison to the existing ones.

To achieve this, one possible experiment is to simulate a phylogeny 
according to the mechanisms proposed by the new diversification model, 
then simulate its alignment and measure how well current inference models perform 
at retrieving the original phylogeny.
If this difference is small enough, current models apparently
suffice.

\verb;pirouette; is an R package that performs such an experiment,
and is built on \verb;babette; [Bilderbeek \& Etienne, 2018], 
which calls BEAST2 [\cite{bouckaert2014beast}]. 
With \verb;pirouette;, one
can easily measure the error made by Bayesian inference in recovering
any given phylogeny, helping us to evaluate the necessity of a new diversification model.
Once verified that the introduction of a new model is indeed beneficial,
it can then be added to BEAST2's existing set of (birth-death and 
coalescent) models via the BEAST2 package system.

@@

\iffalse
The goal of \verb;pirouette; is to measure the inference error BEAST2
makes from a given reconstructed phylogeny, within the context of
a simulation study on a (possibly new) diversification model. Such
a study will typically construct phylogenies for different combinations of the diversification model's parameters, to assess under which scenarios the error made by BEAST2 cannot be ignored anymore. 
Such a study needs to be done on, there need be many replicates.
\fi

The goal of \verb;pirouette; is to measure the inference error made by BEAST2 from a given reconstructed phylogeny, which has been simulated according to a (possibly new) diversification model.
To do that the full procedure would be to simulate a great number of phylogenies for different combinations of the diversification model's parameters, to assess under which scenarios the error made by BEAST2 cannot be ignored anymore.

To be able to explain how \verb;pirouette; measures the inference error
BEAST2 makes, we first give two definitions of what
we mean by a generative model and by an inference model: 
a generative model is a combination of site model, clock model and tree prior,
used to simulate alignments. An inference model is a combination of 
site model, clock model, tree prior, an optional node calibration prior
and setup of the Markov chain Monte Carlo (MCMC) algorithm, 
used to produce a posterior. Node calibration priors, called 'most
recent common ancestor' (MRCA) priors within BEAST2, allow to create a time-calibrated phylogeny, by specifying a normal distribution around an expected time when the ancestors of two or more taxa diversified.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{workflow.png}
  \caption{
    \texttt{pirouette} pipeline. 
    The pipeline starts from a true phylogeny (1a), 
    that is converted to an alignment (2a).
    The experiments are filtered (3a), possibly based on the
    alignment (2a), to determine which experiments are actually run.
    The experiment that are actually performed, use the 
    alignment (2a) to create a BEAST2 posterior of phylogenies (4a),
    which is compared to the true phylogeny (1a) 
    using the nLTT statistic. (5a) shows the nLTT of the true tree in red,
    the nLTT of all individual posterior trees in grey, and the average
    nLTT statistic of all posterior trees in black.
    Taking the differences between the true tree's nLTT and each posterior's
    nLTT results in an error distribution, here displayed as a histogram
    of error values (6a).
    Optionally, a twin phylogeny (1b) can be generated from a true
    phylogeny (1a), which then results in an error distribution (6b) with
    the similar intermediate steps 
  }
  \label{fig:pipeline}
\end{figure}

The pipeline is summarized by the following steps, which will be described in detail below:
\begin{enumerate}
    \item from a given phylogeny an alignment is simulated;
    \item from the alignment an inference model is chosen;
    \item the alignment is used as BEAST2 input to infer a posterior distribution of phylogenies;
    \item posterior phylogenies are compared with the given original phylogeny to estimate the error we make. The higher the match, the lower the error;
\end{enumerate}
The pipeline is visualized in Fig.~\ref{fig:pipeline}.
There is also the option to generate a 'twin tree',
that goes through the same pipeline.

The first step simulates a DNA alignment from the given 
phylogeny (Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 2a).
One can specify a DNA sequence
of any length at the root of the phylogeny, a DNA mutation rate, a
site (i.e. nucleotide substitution) model, 
a clock model, a random number generator (RNG) seed and a location
where the alignment is saved to. This step is relatively fast, but longer
DNA alignments will noticeably slow down the inference step.

The second step selects experiments to be actually run, 
based on the experiments and the alignment (Fig.~\ref{fig:pipeline}, 2a $\rightarrow$ 3a) 
The user can specify whether to use the generative model or select the best model from a set of inference models. 
When selecting the generative model, the site and clock model used in the alignment simulation are used in the inference. Because the given phylogeny (on which the alignment is based)
may have followed any tree prior (i.e. speciation model), the user needs
to specify which tree prior is used in the inference. 
When selecting for the best
model, the alignment is used to find the inference model that has the
highest evidence (i.e. marginal likelihood) from a set of candidate inference models.
The evidence of an inference model is estimated using a nested sampling (NS)
approach, as described in \cite{maturana2018model}. The nested sampling is
performed by \verb;mcbette; [\cite{mcbette}], that calls the 'NS' BEAST2 package. 
Using BEAST2 packages (in a scripted way) can only be done under Linux and Mac; 
inference model selection is not available on Windows.

The third step infers a Bayesian posterior from the simulated 
alignment (Fig.~\ref{fig:pipeline}, 2a $\rightarrow$ 3a $\rightarrow$ 4a),
using the inference model(s) selected in the previous step. The user
can specify the additional parameters needed for the BEAST2 run, which
are the Markov-Chain Monte Carlo (MCMC) setup, 
an optional MRCA prior and an RNG seed \giovanni{This has been already said when defining the inference model. I suggest to define it only once in detail (like here) and just mention it the first time (something like: "An inference model is a combination of site model, clock model, tree prior as well as additional BEAST2 settings." in the description section, around line 98), which will probably look a bit more neat, as it's a bit clearer the parallel between alignment model and inference model.}.
The MCMC setup determines the number of posterior trees sampled.
An MRCA prior allows the inferred phylogeny to have a dated crown age.

@@

An optional step is to generate a 'twin tree' (Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 1b),
that will be analyzed in the same way as the true tree.
The twin tree has the same topology as the given phylogeny, 
yet its branching times are simulated according to 
the tree prior used by BEAST2 to produce a posterior 
\giovanni{is it true? is it Yule or BD? Please check}
\richel{It is Yule}.
The goal of this procedure is to generate a tree 
to use as a control for the main experiment.
We produce the twin tree according to the following procedure.

From the given (true) phylogeny, we maximize a birth-death 
likelihood [\cite{nee1994reconstructed}] to infer the standard 
birth-death rates, namely speciation rate $\lambda$ and extinction rate $\mu$.
We use $\lambda$ and $\mu$ to simulate a new birth-death tree, 
conditioned on having the same number of tips as the original tree. 
We store its branching times. 
We then combine such branching times with the original tree's 
topology to obtain the twin tree.

@@

\verb;pirouette; can be installed easily from CRAN:

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
install.packages("pirouette")
\end{lstlisting}

>>>>

% Affiliations
\usepackage{authblk}
\title{pirouette: Determining the error BEAST2 makes in inferring a phylogeny}
\author[1]{Rich\`el J.C. Bilderbeek}
\author[1]{Giovanni Laudanno}
\author[1]{Rampal S. Etienne}
\affil[1]{Groningen Institute for Evolutionary Life Sciences, University of Groningen, Groningen, The Netherlands}

@@

%comments
\newcommand{\giovanni}[1]{\textcolor{blue}{\textbf{[GL: #1]}}}
\newcommand{\richel}[1]{\textcolor{orange}{\textbf{[RB: #1]}}}
\newcommand{\rampal}[1]{\textcolor{green}{\textbf{[RSE: #1]}}}

@@

  \textbf{1. }
    BEAST2 is a popular Bayesian phylogenetics software tool,
    that takes an alignment of characters (usually nucleotides) and an inference model to create a
    posterior of jointly-estimated phylogenies and model parameter estimates. An important ingredient of BEAST2 is the model for the species tree prior, and the tool can in principle be used for inference of macroevolutionary processes underlying the species tree. However, only models that allow for fast computation of the prior probability are possible in practice. This begs the question how accurate the tree estimation is when the real macroevolutionary processes are different from those assumed in BEAST2. 
  \textbf{2. }
    Here we present \verb;pirouette;, 
    a free, libre and open-source R package that assesses 
    the inference error BEAST2 makes based on a known/true 
    phylogeny, for example generated by a new macroevolutionary diversification model. \\
  \textbf{3. }
    We describe \verb;pirouette;'s usage and the biological scientific
    question it can answer, including full examples. \\
  \textbf{4. }
    Last, we discuss the results obtained by the examples. \\
\end{abstract}

@@

The development of new powerful inference tools, 
such as BEAST [\cite{drummond2007beast}], 
MrBayes [\cite{huelsenbeck2001mrbayes}]
or RevBayes [\cite{hohna2016revbayes}], 
allows to build phylogenetic trees 
from character data (usually nucleotide sequences) extracted from extant organisms.
This has constituted an important step forward 
in our understanding of how species evolve.
Such tools have been increasingly exploited to hypothesize 
and test what the main drivers and modes for diversification are.

Because BEAST performs a Bayesian analysis, 
it does not only need genetic data but also tree priors, 
describing how the diversification occurs, to return posterior phylogenies.
BEAST uses standard tree priors such as Yule or constant-rate birth-death models
\giovanni{integrate this if necessary}.
The recent development of BEAST2 [\cite{bouckaert2014beast}] provides, 
unlike its predecessor, the possibility for third-party users 
to use their own preferred diversification model if an algorithm is provided to calculate the probability of a tree given this model. 
Many such models and their associated probability algorithms have been developed, 
 \giovanni{Not of them are already available for BEAST2 yet, or maybe they will never be}: 
birth-death models that have rates that are 
constant [\cite{nee1994reconstructed}], 
time-dependent [\cite{nee1994reconstructed}, \cite{rabosky2008explosive}], 
or diversity-dependent [\cite{etienne2011diversity}],
or that can shift [\cite{etienne2012conceptual}, \cite{rabosky2014automatic}, \cite{alfaro2009nine}, the SLS model. Laudanno et al., in preparation].
Other birth-death models treat speciation as a process that takes time [\cite{rosindell2010protracted}][\cite{etienne2012prolonging}], or that happens in bursts of simultaneous events [Laudanno et al., in preparation], or that
depends on a trait that has two [\cite{maddison2007estimating}], 
or more [\cite{fitzjohn2012diversitree}] states,
or even allows concealed states [\cite{beaulieu2016detecting}] 
or a combination of all these [\cite{herrera2018detecting}].

For a given phylogeny the parameter values of the diversification model are usually estimated using maximum likelihood.
For such  models, it is a standard practice to assess their performance 
on a controlled environment: phylogenies are simulated with known parameters, and then the estimated parameters are compared with the original parameters.
If the likelihood proves to be effective to describe the process, 
it can potentially be implemented into BEAST2 as a new prior. However, in many cases the computation of this likelihood (the tree prior) is computationally demanding and it is unfeasible to use it in a Bayesian inference framework where it needs to be computed millions to billions of times.
However, if the data are very informative, the influence of the prior is often relatively small, and hence under this assumption it has become common practice to use already existing BEAST2 priors. However, it is unclear whether this assumption holds.
Here we introduce a pipeline to assess this assumption for any given diversification model. Briefly, it consists of the following steps: first simulate a phylogeny 
according to the mechanisms proposed by the new diversification model, 
then simulate a nucleotide alignment and use current inference tools (such as BEAST2) to estimate the phylogeny. If the estimated tree differs only slightly from the original phylogeny, the original assumption is justified and the simple diversification models in the inference tools suffice. If the trees differ markedly, it will be worth the effort and computational burden to develop a species tree module to be used in these tools.
Our pipeline is programmed as an R package and is called 
\verb;pirouette; it is built on \verb;babette; [Bilderbeek \& Etienne, 2018], 
which calls BEAST2 [\cite{bouckaert2014beast}]. 
With \verb;pirouette;, one
can easily measure the error made by Bayesian inference in recovering
any given phylogeny, helping us to evaluate the necessity of a new diversification model.

@@

The goal of \verb;pirouette; is to measure the inference error BEAST2
makes for a given reconstructed phylogeny, within the context of
a simulation study on a (possibly new) diversification model. Such
a study will typically construct phylogenies for different combinations of the diversification model's parameters, to assess under which scenarios the error made by BEAST2 cannot be neglected. 
This usually requires many replicates.

To explain how \verb;pirouette; measures the inference error
BEAST2 makes, we first give two definitions of what
we mean by a generative model and by an inference model: 
a generative model is a combination of site model, clock model and tree prior,
used to simulate DNA alignments. An inference model is a combination of 
site model, clock model, tree prior, an optional node calibration prior
and setup of the Markov chain Monte Carlo (MCMC) algorithm, 
used to produce a posterior distribution of trees and model parameters. Node calibration priors, called 'most
recent common ancestor' (MRCA) priors within BEAST2, allow to create a time-calibrated phylogeny, by specifying a normal distribution around an expected time when the ancestors of two or more taxa diversified.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{workflow.png}
  \caption{
    \texttt{pirouette} pipeline. 
    The pipeline starts from a true phylogeny (1a), 
    that is converted to an alignment (2a). \rampal{this is unclear; the reader does not know what an experiment is, why there are more, what filtering means, etc}
    The experiments are filtered (3a), possibly based on the
    alignment (2a), to determine which experiments are actually run.
    The experiment that are actually performed, use the 
    alignment (2a) to create a BEAST2 posterior of phylogenies (4a),
    which is compared to the true phylogeny (1a) 
    using the nLTT statistic [\cite{janzen2015approximate}]. (5a) shows the nLTT of the true tree in red,
    the nLTT of all individual posterior trees in grey, and the average
    nLTT statistic of all posterior trees in black.
    Taking the differences between the true tree's nLTT and each posterior's
    nLTT results in an error distribution, here displayed as a histogram
    of error values (6a).
    Optionally, a twin phylogeny (1b) can be generated from a true
    phylogeny (1a), which then results in an error distribution (6b) with
    similar intermediate steps 
  }
  \label{fig:pipeline}
\end{figure}

The pipeline is summarized by the following steps, which will be described in detail below:
\begin{enumerate}
    \item for a given phylogeny an alignment is simulated for a known 'alignment' model;
    \item for this alignment, model selection is performed to choose the best-fitting alignment model in inference (which may be different from the generative alignment model);
    \item the alignment is then used as BEAST2 input to infer a posterior distribution of phylogenies;
    \item posterior phylogenies are compared with the given original phylogeny to estimate the error made;
\end{enumerate}
The pipeline is visualized in Fig.~\ref{fig:pipeline}.
There is also the option to generate a 'twin tree',
that goes through the same pipeline. The utility of this twin tree will be explained below.

The first step simulates a DNA alignment from a given 
phylogeny (Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 2a).
One can specify a DNA sequence
of any length at the root of the phylogeny, a DNA mutation rate, a
site (i.e. nucleotide substitution) model, 
a clock model, a random number generator (RNG) seed and a location
where the alignment is saved to. This step is relatively fast, but longer
DNA alignments will noticeably slow down the inference step.

In the second step selects (Fig.~\ref{fig:pipeline}, 2a $\rightarrow$ 3a), the user can specify whether to use the generative model as the inference model or to select the best model from a set of inference models. 
When selecting the generative model, the site and clock model used in the alignment simulation are used in the inference. Furthermore, because the given phylogeny (on which the alignment is based)
may have followed any tree prior (i.e. speciation model), the user needs
to specify which tree prior is used in the inference. 
When selecting for the best
model, the alignment is used to find the inference model that has the
highest evidence (i.e. marginal likelihood) from a set of candidate inference models.
The evidence of an inference model is estimated using a nested sampling (NS)
approach, as described in \cite{maturana2018model}. The nested sampling is
performed by \verb;mcbette; [\cite{mcbette}], that calls the 'NS' BEAST2 package. 
Using BEAST2 packages (in a scripted way) can only be done under Linux and Mac; 
inference model selection is not available on Windows.\rampal{clarify this. One can use mcbette on a browser, right?}

The third step infers the actual Bayesian posterior from the simulated 
alignment (Fig.~\ref{fig:pipeline}, 2a $\rightarrow$ 3a $\rightarrow$ 4a),
using the inference model(s) selected in the previous step. The user
can specify the additional parameters needed for the BEAST2 run, which
are the Markov-Chain Monte Carlo (MCMC) setup, 
an optional MRCA prior and an RNG seed \giovanni{This has been already said when defining the inference model. I suggest to define it only once in detail (like here) and just mention it the first time (something like: "An inference model is a combination of site model, clock model, tree prior as well as additional BEAST2 settings." in the description section, around line 98), which will probably look a bit more neat, as it's a bit clearer the parallel between alignment model and inference model.}.
The MCMC setup determines the number of posterior trees sampled.
An MRCA prior allows the inferred phylogeny to have a dated crown age.

@@

An optional step is to generate a 'twin tree' (Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 1b),
that will be analyzed in the same way as the true tree.
The twin tree has the same topology as the given phylogeny, 
yet its branching times are simulated according to 
the tree prior used by BEAST2 to produce a posterior 
\giovanni{is it true? is it Yule or BD? Please check}
\richel{It is Yule}.
\rampal{I think it's better to use BD for more generality. Yule is too limited in the trees it can produce.}
The goal of this procedure is to generate a tree 
to use as a control for the main experiment.
We produce the twin tree according to the following procedure.

From the given (true) phylogeny, we maximize a birth-death 
likelihood [\cite{nee1994reconstructed}] to infer the standard 
birth-death rates, namely speciation rate $\lambda$ and extinction rate $\mu$.
We use $\lambda$ and $\mu$ to simulate a new birth-death tree, 
conditioned on having the same number of tips as the original tree, we store its branching times, and then  combine the branching times with the original tree's 
topology to obtain the twin tree.

@@

\verb;pirouette; can be installed easily from CRAN:
\rampal{be careful here; it is not on CRAN yet. Better to say it can be installed from GitHub and will be made available on CRAN)}
\begin{lstlisting}[language=R, floatplacement=H, frame=single]
install.packages("pirouette")
\end{lstlisting}
